A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-24 text/manuscript.Rmd
update fig 3 & 5 legends, typos, survival text

**Figure 1: Molecular pathway profiles of tumor samples with one or more
mutations.** Each of 377 selected Reactome pathways (columns) is classified as
disrupted if one or more genes is mutated in the tumor sample (rows) where red
and grey represents the pathway is mutated and not, respectively. Tumors are
grouped by tissue of origin using standardized abbreviations from the TCGA
project.

@@

```{r, child='../code/figure2.Rmd'}
```

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types (with some notable
exceptions) were not heavily biased in one class, and all well-represented
cancer types had tumors in every class (see **figure 3A** and full tumor
profiles in **supplementary figure S1**, see also interactive media from
**supplementary file 1 (html)**), suggesting that, in principle at least, our
pathway-disruptions identify clusters of molecular pathology largely independent
of tissue-of-origin. As an example of one type of cancer that does have a biased
pathway profile, pancreatic adenocarcinoma (PAAD) was predominantly found in
class 8 (**figure 3A** and **supplementary figure S1**). But even PAAD comprises
tumors from the nine remaining classes, meaning that patients suffering these
tumors have potentially different underlying molecular pathologies.

@@

**Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
a) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
Pancreatic adenocarcinoma. b) Projection of breast cancer subtypes onto the
UMAP. c) Projection of BRCA1/2 somatic mutation onto the UMAP. d) Projection of
tumor stage onto the UMAP, regardless of cancer type. e) Projection of
metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
metastatic tumors.

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2C**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption classes (**figure 3D**), suggesting that stage is not an
determining factor in pathway disruption classes.

@@

**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
a) Waterfall plot; columns represent tumors. Each row represents a pathway. The
pathways are ranked such that the pathway with the most disruptions across all
7,607 tumors is the first row _etc._ and tumors are ordered to prioritize the
top pathway disruptions successively. The top 100 most mutated pathways are
shown. Red signifies pathway disrupted pathways, grey signifies not disrupted.
b) Heatmap shows the relative enrichment of each pathway (columns) within each
numbered cluster (rows). Effect size is displayed as colors (orange, enrichment;
purple, depletion). c) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size >= 0.30 in each
cluster) within each numbered cluster (rows).

@@

```{r child='../text/fig_legend.Rmd'}
```

>>>>

**Figure 1: Molecular pathway profiles of tumor samples with one or more
mutations.** Each of 377 selected Reactome pathways (rows) is classified as
disrupted if one or more genes is mutated in the tumor sample (columns) where
red and grey represents the pathway is mutated and not, respectively. Tumors are
grouped by tissue of origin using standardized abbreviations from the TCGA
project.

@@

#### Figure 2
```{r, child='../code/figure2.Rmd'}
```

**Figure 2: Clustering of tumor samples** Different rotational perspectives of
the same MCA-based UMAP projection in three-dimensional space. Each dot
corresponds to a tumor sample. The same colors indicate the tumor's cluster
identity throughout this manuscript. 1=yellow, 2=lilac, 3=orange, 4=light blue,
5=crimson, 6=beige, 7=grey, 8=dark green, 9=pink, 10=dark blue.

@@

Having defined tumors in terms of their pathway disruption profile, we sought to
understand whether different cancer types segregated into one or more
predominant classes. To our surprise, most cancer types were not heavily biased
in one class, and all well-represented cancer types had tumors in every class
(see **figure 3A** and full tumor profiles in **supplementary figure S1**, see
also interactive media from **supplementary file 1 (html)**), suggesting that,
in principle at least, our pathway-disruptions identify clusters of molecular
pathology largely independent of tissue-of-origin. As an example of one type of
cancer that does have a biased pathway profile, pancreatic adenocarcinoma (PAAD)
was predominantly found in class 8 (**figure 3A** and **supplementary figure
S1**). But even PAAD comprises tumors from the nine remaining classes, meaning
that patients suffering these tumors have potentially different underlying
molecular pathologies.

@@

**Figure 3: Pathway-based classification is independent of tissue-of-origin features.** 
A) Sankey plot showing correspondence between cancer type and cluster identity.
B) Projection of cancer type identity onto UMAP. Abbrevs: BRCA = Breast cancer,
COAD = Colorectal adenocarcinoma, HNSC = Head and neck squamous cell, PAAD =
Pancreatic adenocarcinoma. C) Projection of breast cancer subtypes onto the
UMAP. D) Projection of BRCA1/2 somatic mutation onto the UMAP. E) Projection of
tumor stage onto the UMAP, regardless of cancer type. E) Projection of
metastatic status onto the UMAP. Abbrevs: M0 = non-metastatic tumors, M1 =
metastatic tumors.

@@

Tumor staging is based on physico-pathological criteria, including tumor
diameter, which can vary greatly in importance between different tissues. Stage
is used clinically as a proxy for advancement toward a more deadly state and
metastasis. Given these assumptions, it is possible that more advanced tumors
have common pathway disruption profiles, and the UMAP which features a series of
lobe-like structures on a common backbone of tumor samples (**figure 2**) could
in principle also reflect progression through a series of stages. In support of
this hypothesis, the backbone starts with a cluster of tumors (class 1) that has
the fewest point mutations and culminates in a cluster (class 10) which has
nearly every pathway disrupted (**figure 4**). However, outside of class 10 we
don't observe an obvious trend in the overall mutation burden across the
backbone of the UMAP (**figure 1H**). Nonetheless, to test the hypothesis that
the molecular-pathway disruption clusters represent advancement through stages,
we projected staging data onto the UMAP. Similar to tissue of origin and other
categories of tumor, we did not observe any bias among the stages to specific
pathway disruption classes (**figure 3D**), suggesting that stage is not an
determining factor in pathway disruption classes.

@@

**Figure 5: Pan-cancer enrichment of pathway disruptions.** 
A) Heatmap shows the relative enrichment of each pathway (rows) within each
numbered cluster (columns). Effect size is displayed as colors representing
percent enrichment. B) Proportion of mutated samples in the each significant
pathway (columns; union set of pathways with effect size ≥ 0.30 in each
cluster) within each numbered cluster (rows).

@@

### 2020-08-24 text/survival.Rmd
update fig 3 & 5 legends, typos, survival text

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are profoundly limited by
confounding factors of age and stage at diagnosis, sex, ethnicity, and
tissue-specific disease progression, to name the obvious ones. To explore these
ideas, we used Bayesian inference to test models of survival using public
longevity data from the CDC, accounting for age, gender and ethnicity. Our model
estimates the effect of cancer type and class-specific cancer effects
independently, resulting in a cancer and class specific estimate of the aging
rate multiplier.

To understand the model, we show in **Figure 7A** simulated survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated without the confounding aspect of
age.

@@

**Figure 7: Comparison of different cancers, simulated and actual results
results.** A) To understand the risk model, we plot survival of four simulated
cohorts. Compare survival of a group consisting of randomly selected 30 year old
women (light blue) to an equal sized group of 30 year old women with cancer, k =
2 (light orange). To understand how the mix of ages alters the survival curve,
we compare a randomly selected group of people with ages uniformly distributed
from 30 to 70 (dark blue) and a randomly selected group with uniformly
distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
selected group with correct age distribution for the given cancer (turqoise) to
the model predicted survival for the given average k (blue) and the actual
survival for the patients in our dataset (orange). As can
be seen, Stomach Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD)
are all very deadly. The modeled results fit the general trend for the actual
patients.  Thyroid Cancer (THCA) on the other hand has very little if any effect
on our estimates of life expectancy, in fact there may be a slight benefit which
may be associated with a bias in diagnosis towards patients who are more health
conscious, or higher socioeconomic status, with more access to care compared to
random population members.

>>>>

---
output:
  word_document: default
  html_document: default
---
If our molecular pathway disruption classes represent distinct biological states
apart from tissue of origin, they may have different prognoses within cancer
types or across all cancers. These analyses are limited by confounding factors
of age and stage at diagnosis, sex, ethnicity, and tissue-specific disease
progression. To explore these ideas, we used Bayesian inference to test models
of survival using public longevity data from the CDC, accounting for age, gender
and ethnicity. Our model estimates the effect of cancer type and
cluster-specific cancer effects independently, resulting in a cancer and
cluster-specific estimate of the aging rate multiplier, k.

To understand the model, we show in **Figure 7A** simulated survival time curves
for four related groups. We start with a cohort of women aged 30 and show their
expected survival (light blue). By setting their cancer rate multiplier to $k=2$
we can simulate the effect of a moderately deadly cancer (yellow). Compare this
to a randomly selected group of women ranging in age from 30 to 70 years old
(dark blue). Immediately the survival curve changes due to the mixture of ages,
without any malignancy. Adding a $k=2$ malignancy further reduces expected
lifespan (red). We can see that for different tissues, simply due to the change
in distribution of age-at-diagnosis, we should expect equally deadly cancers to
have *different* survival curves. The model takes this into account and the
aggressiveness of the cancer can be estimated accurately without the confounding
effect of age.

@@

**Figure 7: Comparison of different cancers, simulated and actual results
results.** To illustrate the risk model, we plot survival of four simulated
cohorts. A) Compare survival of a group consisting of randomly selected 30 year old
women (light blue) to an equal sized group of 30 year old women with cancer, k =
2 (light orange). To illustrate how the mix of ages alters the survival curve,
we compare a randomly selected group of people with ages uniformly distributed
from 30 to 70 (dark blue) and a randomly selected group with uniformly
distributed ages and cancer k = 2 (dark orange). B,C,D,E) We compare a randomly
selected group with correct age distribution for the given cancer (turqoise) to
the model predicted survival for the given average k (blue) and the actual
survival for the patients in our dataset (orange). As can be seen, Stomach
Cancer (STAD), Glioblastoma (GBM), and Pancreatic Cancer (PAAD) are all very
deadly. The modeled results fit the general trend for the actual patients.
Thyroid Cancer (THCA) on the other hand has very little if any effect on our
estimates of life expectancy, in fact there may be a slight benefit which may be
associated with a bias in diagnosis towards patients who are more health
conscious, or higher socioeconomic status, with more access to care compared to
random population members.
